HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-04-25 docs/HLL_exp_bad/main.tex
better attack analysis

\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data stream's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. In this cautionary note, we show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm, an adversary who has some flexibility in choosing what inputs will be recorded can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. 
\end{abstract}
}

@@

In scenario S2, the adversary needs to craft two distinct candidates for $x$ on average to obtain a one bit at the first position in the $\ell$ rightmost bits of $h(x)$, thus successfully performing the attack for a single insertion. Assuming the adversary has $t$ bits of flexibility in its choice of inputs, the adversary can expect to ``hide''  2$^{t-1}$ distinct items without increasing the HLL cardinality estimate at all.

@@

Note that the above analysis made the assumption that every bucket count $c_i$ was already at least 1. When some bucket counts are 0, the above analysis needs to be modified in order to avoid touching those buckets altogether. The modified analysis is quite straightforward. Suppose $r \le 2^n-1$ out of $2^n$ buckets have a count of zero. Then the cost of crafting an input $x$ becomes $(1-(2H_c)^{-1})^{-1} \cdot (1- \frac{r}{2^n})^{-1}$ trials, while the total number of items the adversary can expect to insert becomes $(1-(2H_c)^{-1}) \cdot (1- \frac{r}{2^n})\cdot 2^t$.

We assumed so far that there are already values coming from other honest users' data in the HLL. We now remove this assumption. Unfortunately, the adversary in scenario S2 has no way to detect that he is attacking an empty HLL and cannot adapt his strategy, so all buckets hit will update their estimator once to $2^{0+1}=2$ items, leading to a final cardinality approximation of $\alpha2^{n+1}$ regardless of the number of items the adversary inserts. However, the adversary in S4 is able to detect that all buckets are empty and adapt his attack accordingly. The adversary can target and fill one bucket only. This requires fixing $n+1$ bits of $h(x)$ for each input $x$, allowing the adversary to ``hide'' $2^{t-n-1}$ elements while keeping the HLL cardinality estimate to 1. Other work/cardinality estimate trade-offs are of course possible.

\subsection{Experimental Results}
There exist many implementations of the HyperLogLog algorithm available online. Among the repositories collecting more than 350 stars on github, we can cite \cite{clahll} and \cite{datasketch}. It is also featured in many frameworks, including Redis \cite{redis}, Google's BigQuery \cite{bigquery} and several products of Apache such as Spark \cite{spahll} \cite{spahll2} and Druid \cite{druhll}. It is interesting to point out that they all use a constant number of buckets and a fixed hash function, \textit{which is public or should be public according to Kerckhoffs's principle,} thus are potentially vulnerable to cardinality estimate manipulation when put in an adversarial setting.\\

As proof-of-concept, we implement our attack against the HLL algorithm from \cite{clahll}. The code is available at \href{https://github.com/PizzaWhisperer/HLLVuln}{\textit{\url{https://github.com/PizzaWhisperer/HLLVuln}}}. Our items are 4 character long ASCII strings \textit{(7'311'616 distinct items)} and we choose $h$ to be the 32 bits Murmur3 hash function from \cite{murmur3code}. We create an HLL sketch using 256 buckets and initialize it with 1'000 random items representing an honest user's data.

We challenge the adversary to pick the largest possible number of items to add to the sketch while trying to keep the cardinality estimate as low as possible, either in the S2 or the S4 scenario. In a second time, in order to get results that can be compared to the work of \cite{hllvuln}, we mount our attack in their more restrictive setting: instead of picking from the universe of possible strings, the adversary has to choose from a set of 250'000 random items that are given to him. Finally , for completeness, we also run the attacks against empty sketches. All attacks are ran 30 times and the outcomes are reported in Tab. \ref{table:tab1} and Tab. \ref{table:tab2}, which detail the estimated cardinality at the beginning of the attack, the number of items added by the adversary and the approximation of the cardinality after adding the adversary's items.

>>>>

\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data stream's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. In this cautionary note, we show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm and has some flexibility in choosing what inputs will be recorded, it can manipulate the cardinality estimate to be exponentially smaller than the true cardinality. 
\end{abstract}
}

@@

In scenario S2, the adversary needs to craft two distinct candidates for $x$ on average to obtain a one bit at the first position in the $\ell$ rightmost bits of $h(x)$, thus successfully performing the attack for a single insertion. Assuming the adversary has $t$ bits of flexibility in its choice of inputs, the adversary can expect to ``hide''  2$^{t-1}$ distinct items without increasing the HLL cardinality estimate at all, under the assumption that all buckets in the HLL sketch already hold a value $c_i ge 1$. On the other hand, if some buckets are empty (but the adversary does not know which ones in the blind setting of S2), then we may expect a moderate increase in the HLL cardinality estimator. In the worst case, when all buckets are initially empty, the cardinality estimate will increase from 0 to at most $\alpha2^{n+1}$ (regardless of the number of items the adversary inserts).

@@

Note that the above analysis for scenario S4 made the assumption that every bucket count $c_i$ was already at least 1. When some bucket counts are 0, the above analysis needs to be modified in order to avoid touching those buckets altogether. The modified analysis is quite straightforward. Suppose $r \le 2^n-1$ out of $2^n$ buckets have a count of zero. Then the cost of crafting an input $x$ becomes $(1-(2H_c)^{-1})^{-1} \cdot (1- \frac{r}{2^n})^{-1}$ trials, while the total number of items the adversary can expect to insert without increasing the HLL cardinality estimate becomes $(1-(2H_c)^{-1}) \cdot (1- \frac{r}{2^n})\cdot 2^t$. Of course, the adversary can relax its attack and allow some of the empty buckets to be hit. This increases the insertion rate at the cost of an increased final cardinality estimate.

We assumed so far that there are already values coming from other honest users' data in the HLL. We now remove this assumption. The blind adversary in scenario S2 has no way to detect that he is attacking an empty HLL and cannot adapt his strategy, so all buckets that are hit will update their estimator once to $2^{0+1}=2$ items, leading to a final cardinality approximation of $\alpha2^{n+1}$ regardless of the number of items the adversary inserts. However, the adversary in S4 is able to detect that all buckets are empty and adapt his attack accordingly. The adversary can now target and fill one bucket only. This requires fixing $n+1$ bits of $h(x)$ for each input $x$, allowing the adversary to ``hide'' $2^{t-n-1}$ elements while keeping the HLL cardinality estimate to 1. Other work/cardinality-estimate trade-offs are of course possible.

In both scenarios S2 and S4, we see that an adversary who has $t$ bits of flexibility in its inputs can expect to insert exponentially (in $t$) many inputs values into the HLL sketch without increasing the cardinality estimate (or whilst increasing the cardinality estimate slightly in the cases where the HLL sketch has empty buckets in S2 or where the HLL sketch is empty to begin with in S2 and S4). The result is largely independent of the HLL parameters, but does depend on the extent to which the HLL is already filled, via a factor $1-(2H_c)^{-1}$ and the number of empty buckets $r$, via a factor $1- \frac{r}{2^n}$.

\subsection{Experimental Results}
Many different implementations of HLL are available online. Among the repositories collecting more than 350 stars on GitHub, we can cite \cite{clahll} and \cite{datasketch}. HLL is also featured in many frameworks, including Redis \cite{redis}, Google's BigQuery \cite{bigquery} and several products of Apache such as Spark \cite{spahll,spahll2} and Druid \cite{druhll}. It is interesting to point out that all these implementations use a constant number of buckets and a fixed hash function. Thus they are vulnerable to our attacks in scenarios S2 and S4.

As a proof-of-concept, we implemented our attacks against the HLL implementation from \cite{clahll}. The attack code is available at \href{https://github.com/PizzaWhisperer/HLLVuln}{\textit{\url{https://github.com/PizzaWhisperer/HLLVuln}}}. For simplicity, our items are 4-character long ASCII strings, giving us up to 7,311,616 distinct items that can be inserted. We choose $h$ to be the 32-bit Murmur3 hash function from \cite{murmur3code}. \kenny{Why do we get to choose $h$? Isn't it fixed by the specific implementation we are attacking?} We create an HLL sketch using $2^n = 256$ buckets and initialize it with 1,000 random items representing honest users' data.

We then challenge the adversary to pick the largest possible number of items to add to the sketch from our set of. while trying to keep the cardinality estimate as low as possible, either in the S2 or the S4 scenario. In a second time, in order to get results that can be compared to the work of \cite{hllvuln}, we mount our attack in their more restrictive setting: instead of picking from the universe of possible strings, the adversary has to choose from a set of 250'000 random items that are given to him. Finally , for completeness, we also run the attacks against empty sketches. All attacks are ran 30 times and the outcomes are reported in Tab. \ref{table:tab1} and Tab. \ref{table:tab2}, which detail the estimated cardinality at the beginning of the attack, the number of items added by the adversary and the approximation of the cardinality after adding the adversary's items.
